\toolname: file-based localization of numerical perturbations in data analysis pipelines

### 2020-04-10 paper.tex
Fixed notes format
Commented on @gkiar's comments.

%%%% Commands %%%%
\newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
\newcommand{\note}[2]{\color{blue}Note: #1\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{blue}\textbf{From Tristan:}#1\color{black}}
\newcommand{\english}[1]{\uwave{#1}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

@@

Numerical stability is believed to play a central role in the differences
observed in data analysis pipelines across computational infrastructures. \note{Greg}{cite or soften claim}
Variability resulting from operating system upgrades, parallelization, or
even minor data perturbations may originate in errors that accumulate in
numerical computations with important consequences on the reproducibility
of analytical results. However, it remains challenging to identify the
exact elements in a complex pipeline that are prone to numerical instability,
leaving pipeline developers and users with the daunting task to manually
compare results obtained across different conditions.

@@

\newpage

@@

\toolname identifies the components in a pipeline, at the resolution level of a system process, that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each pipeline execution, where nodes
represent application processes and files, and edges represent read and
write file accesses (Figure~\ref{fig:provenance-graph}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Figure~\ref{fig:provenance-dag}). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Figure~\ref{fig:labeled-dag}).
 The resulting labeled graph identifies the
non-reproducible processes in the pipeline. \note{Greg}{clarify why many processes don't appear in the graph as shown}

@@

We did not observe any within-OS difference in PreFreeSurfer. In
FreeSurfer, we identified 2 processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\texttt{mri\_segreg}, and cortical surface curvature estimations with
\texttt{mris\_curvature}. Fixing the random seed used in FreeSurfer removed
these differences. \note{Greg}{can you make some statement on the magnitude of these differences?}

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\begin{minipage}{\linewidth}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. \note{Greg}{colorbar doesn't match the numbers: it's from 0-1 but the numbers are 
  subject counts. I don't really have a preference on which, but harmonizing them to be the same unit should happen.}
        }
  \label{fig:pfs_heatmap}
\end{minipage}

@@

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes.\note{Greg}{This is very hard to see from the figure because the
luminosity of the two colormaps are so different - the contrast moving from orange to grayscale is difficult. Perhaps use two colors, green and purple for instance, with the same intensity to see if this is more easily seen?} 
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation and amplification of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that many 
structures have Dice coefficients below 0.9 for some subjects. \note{Greg}{awkward "many" followed by "some". reword?}
\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png} 
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template. An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here}.
} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
  \hspace*{-1cm}
  \includegraphics[width=1.2\linewidth]{figures/dice_regions.pdf}
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20).
\note{Greg}{I don't think the legend is important here. Focus on having values legible and removing those which
don't really illustrate the "point".}}
    \label{fig:scatter_plot}
  \end{figure}

@@

\subsection{Key findings}
\note{Greg}{feel free to ignore, but a peeve of mine is one heading directly following another. Is there some
kind of high-level summary you could put above this, or just omit the key-findings label?}
Linear and non-linear registration with FSL were found to
frequently lead to differences between results obtained with different
operating systems. This does not come as a surprise given the instabilities
associated with these processes. It also corroborates our previous findings
in~\cite{Glatard2015}, where fMRI pre-processing with FSL was found to vary across operating systems 
starting from the motion
correction step, a step that uses FSL's \flirt tool internally. It
would be relevant to investigate if the observed instability of
registration processes generalizes to other toolkits, or if it remains specific
to FSL. In view of the effect of small data perturbations in a variety of
toolboxes and processes, such as cortical surface extraction using
FreeSurfer and CIVET~\cite{Lewis2017-ll} or connectome estimation using
Dipy~\cite{kiar2019comparing}, it is probable that this observation
generalizes widely across toolboxes and requires a deeper investigation of
the stability of linear and non-linear registration.

>>>>

%%%% Commands %%%%
\newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
\newcommand{\note}[2]{\color{blue}Note from \textsc{#1}: #2\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\english}[1]{\uwave{#1}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

@@

Computational infrastructures 
are known to impact data analysis pipelines, but it remains difficult to 
\tristan{I will finish this.}
Numerical stability may play a central role in the differences
observed in data analysis pipelines across computational infrastructures. \note{Greg}{cite or soften claim}
Variability resulting from operating system upgrades, parallelization, or
even minor data perturbations may originate in errors that accumulate in
numerical computations with important consequences on the reproducibility
of analytical results. However, it remains challenging to identify the
exact elements in a complex pipeline that are prone to numerical instability,
leaving pipeline developers and users with the daunting task to manually
compare results obtained across different conditions.

@@

@@

\toolname identifies the components in a pipeline, at the resolution level of a system process, that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each pipeline execution, where nodes
represent application processes and files, and edges represent read and
write file accesses (Figure~\ref{fig:provenance-graph}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Figure~\ref{fig:provenance-dag}). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Figure~\ref{fig:labeled-dag}).
 The resulting labeled graph identifies the
non-reproducible processes in the pipeline. \note{Greg}{clarify why many processes don't appear in the graph as shown} \tristan{Ali, 
do you pre-process the reprozip trace to remove some processes, or does the figure show all the processes in this trace?}

@@

We did not observe any within-OS difference in PreFreeSurfer. In
FreeSurfer, we identified 2 processes leading to within-OS differences due
to the use of pseudo-random numbers: image registration with
\texttt{mri\_segreg}, and cortical surface curvature estimations with
\texttt{mris\_curvature}. Fixing the random seed used in FreeSurfer removed
these differences. \note{Greg}{can you make some statement on the magnitude of these differences?} \tristan{Ali, that's for you.}

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\begin{minipage}{\linewidth}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. \note{Greg}{colorbar doesn't match the numbers: it's from 0-1 but the numbers are 
  subject counts. I don't really have a preference on which, but harmonizing them to be the same unit should happen.} \tristan{
    Ali, it looks like this figure is still unclear, and that other people may still misunderstand it. To better 
    show that the number is a number of subjects, I'd suggest to replace number "x" with "(n=x)",
    write it in smaller font size, and at the bottom-right of the cell instead of in the middle. 
  }
        }
  \label{fig:pfs_heatmap}
\end{minipage}

@@

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes.\note{Greg}{This is very hard to see from the figure because the
luminosity of the two colormaps are so different - the contrast moving from orange to grayscale is difficult. Perhaps use two colors, green and purple for instance, with the same intensity to see if this is more easily seen?} 
\tristan{Ali, I can't find the script to generate this figure in the repo. Could you fix the colors and contrast as per Greg's suggestion?}
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation and amplification of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that many 
structures have Dice coefficients below 0.9 for some subjects. \note{Greg}{awkward "many" followed by "some". reword?}
\tristan{Will do.}
\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png} 
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template. An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here}.
} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure*}
  \hspace*{-1cm}
  \includegraphics[width=1.2\linewidth]{figures/dice_regions.pdf}
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20).
\note{Greg}{I don't think the legend is important here. Focus on having values legible and removing those which
don't really illustrate the "point".}}
    \label{fig:scatter_plot}
  \end{figure*}

@@

\subsection{Key findings}
\note{Greg}{feel free to ignore, but a peeve of mine is one heading directly following another. Is there some
kind of high-level summary you could put above this, or just omit the key-findings label?} \tristan{will do}
Linear and non-linear registration with FSL were found to
frequently lead to differences between results obtained with different
operating systems. This does not come as a surprise given the instabilities
associated with these processes. It also corroborates our previous findings
in~\cite{Glatard2015}, where fMRI pre-processing with FSL was found to vary across operating systems 
starting from the motion
correction step, a step that uses FSL's \flirt tool internally. It
would be relevant to investigate if the observed instability of
registration processes generalizes to other toolkits, or if it remains specific
to FSL. In view of the effect of small data perturbations in a variety of
toolboxes and processes, such as cortical surface extraction using
FreeSurfer and CIVET~\cite{Lewis2017-ll} or connectome estimation using
Dipy~\cite{kiar2019comparing}, it is probable that this observation
generalizes widely across toolboxes and requires a deeper investigation of
the stability of linear and non-linear registration.
